---
title: "Homework 1: Using OLS Regression to Predict Median House Values in Philadelphia"
authors: Chloe Alimurong, Sophia Meyers & Evelyn Zhong
date: October 5, 2026
output: rmdformats::readthedown
---
## Make sure you have results paragraph

# Introduction
## a.
In this analysis, we will assess the relationship between median house values in Philadelphia and variables such as # Households Living in Poverty, % with Bachelor's Degree or Higher, % Vacant Houses and % Single House Units. 
## b. 
The predictors in our analysis describe different aspects of housing that could potentially influence house values and prices. The number of households in poverty and with a bachelors degree, to some degree, explain the economic status of the demographic's housing market. Those living in poverty may not be able to afford houses with higher price tags and those with a bachelors degree may have an income that could afford houses with higher price tags. These population factors indirectly relate to house value, thus why we chose them to predict in our analysis. For unit-specific predictors like vacant houses and single house units, these two signal environmental factors that could influence house value. For vacant houses, we expect that the neighborhood may not be as desirable or may not have high economic activity, making it less desirable to live, which could decrease house values. The percentage of single house units may also highlight the neighborhood's population. In Philadelphia, it is rare to see single-family detached homes. The common house type are multi-family style houses. So for those living in single-family homes, it indicates the desire to live at longer terms and farther from the city center. These types of homes may single out those with higher incomes to afford a single unit house within a city, which may mean higher house value. All of these predictors indicate something about the housing market population and environment indirectly, making them good examples of variables to predict median house value within Philadelphia.

```{r}
#| label: setup
#| warning: false
#| message: false 
# note to self: make sure all libraries are in use or necessary (since tidyverse has a lot)...

library(sf)
library(Hmisc)
library(DAAG)
library(car)
library(MASS)
library(rsq)
library(tidyverse)
library(caret)
library(dplyr)
library(ggplot2)
library(tidyr)
library(patchwork)
```

# Methods
## a. Data Cleaning
The original Philadelphia Block Group dataset had 1816 observations and was cleaned in order to achieve a dataset with 1720 observations. The following items were removed; Block groups where population < 40, Block groups where there are no housing units, Block groups where the median house value is lower than 10,000, One North Philadelphia block group which had a very high median house value (over 800,000), and a very low median household income (less than $8,000). The purpose of their removal was to remove any outliers, no data, and insufficient sample size from our analysis.


## b. Exploratory Analysis
i. In this section, we will evaluate the summary statistics and distributions of the dependent variable and predictors.
```{r}
#| label: summary statistics & examining the distribution of the dependent variable and predictors

#regdata <- read.csv("~/GitHub/MUSA-5000-portfolio/HW1/RegressionData.csv")
regdata <- read.csv("RegressionData.csv")
#names(regdata) # gets the names of each variable.

regdata <- regdata %>% mutate(across(where(is.character), as.numeric)) # some of them were returning a string, we need a number to run summary statistics.

vars <- c("MEDHVAL","PCTBACHMOR", "NBELPOV100", "PCTVACANT", "PCTSINGLES")

sum_table <- data.frame(
  Variable = c("Median House Value",
                        "# Households Living in Poverty",
                        "% with Bachelor's Degree or Higher",
                        "% Vacant Houses",
                        "% Single House Units"),
  Mean = sapply(regdata[vars], mean, na.rm = TRUE), #creates column of mean values for each variable, ignoring null values.
  SD = sapply(regdata[vars], sd, na.rm = TRUE) #creates column of sd values for each variable, ignoring null values.
)

sum_table

#running hist on all of the variables we defined above
for (v in vars){
  hist(regdata[[v]],
       main = paste("Distribution of ", v),
       xlab =v,
       col = "lightblue",
       border = "darkblue")
}

```
ii. In our exploratory data analysis, we will examine the correlations between the predictors. A correlation is , and provide the formula for the sample correlation coefficient r. Also mention the possible range of r values, and what correlation of 0 means.

```{r}
#| label: log transformations and its effect on the distribution

regdata <- regdata %>%
  mutate(LNMEDHVAL    = log(MEDHVAL),
         LNPCTBACHMOR = log(1 + PCTBACHMOR),
         LNBELPOV100 = log(1 + NBELPOV100),
         LNPCTVACANT  = log(1 + PCTVACANT),
         LNPCTSINGLES = log(1 + PCTSINGLES))

hist(regdata$MEDHVAL,     main = "MEDHVAL",     xlab = "MEDHVAL",col = "lightblue",border = "darkblue")
hist(regdata$LNMEDHVAL,   main = "LNMEDHVAL",   xlab = "LNMEDHVAL",col = "lightblue",border = "darkblue")


hist(regdata$PCTBACHMOR,  main = "PCTBACHMOR",  xlab = "PCTBACHMOR",col = "lightblue",border = "darkblue")
hist(regdata$LNPCTBACHMOR, main = "LNPCTBACHMOR", xlab = "LNPCTBACHMOR",col = "lightblue",border = "darkblue")


hist(regdata$NBELPOV100,  main = "NBELPOV100",  xlab = "NBELPOV100",col = "lightblue",border = "darkblue")
hist(regdata$LNBELPOV100, main = "LNBELPOV100", xlab = "LNBELPOV100",col = "lightblue",border = "darkblue")

hist(regdata$PCTVACANT,  main = "PCTVACANT",  xlab = "PCTVACANT",col = "lightblue",border = "darkblue")
hist(regdata$LNPCTVACANT, main = "LNPCTVACANT", xlab = "LNPCTVACANT",col = "lightblue",border = "darkblue")

hist(regdata$PCTSINGLES,  main = "PCTSINGLES",  xlab = "PCTSINGLES",col = "lightblue",border = "darkblue")
hist(regdata$LNPCTSINGLES, main = "LNPCTSINGLES", xlab = "LNPCTSINGLES",col = "lightblue",border = "darkblue")

```


## b.
```{r}
#| label: Scatter plot for each predictor
#| message: false
#| warning: false

regdata %>%
  dplyr::select(LNMEDHVAL, PCTBACHMOR, LNBELPOV100, PCTVACANT, PCTSINGLES) %>%
  pivot_longer(-LNMEDHVAL, names_to = "Predictor", values_to = "Value") %>%
  ggplot(aes(x = Value, y = LNMEDHVAL)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm", se = FALSE, color = "purple") +
  facet_wrap(~Predictor, scales = "free_x") +
  labs(x = NULL, y = "LNMEDHVAL")

```


## c.
```{r}
#| label: Pearson correlation for each predictor
#note to self: make this a matrix with only the correct # of values

cor(regdata$PCTVACANT, regdata$MEDHVAL, method = "pearson")
cor(regdata$PCTSINGLES, regdata$MEDHVAL, method = "pearson")
cor(regdata$PCTBACHMOR, regdata$MEDHVAL, method = "pearson")
cor(regdata$NBELPOV100, regdata$MEDHVAL, method = "pearson")

pearson <- rcorr(as.matrix(regdata), type="pearson")                            
pearson$r 

```
## d.
```{r}
#| label: Chloropleth maps of each variable using the sf (simple features) and ggplot packages
#| note: readOGR has been officially retired as a part of the rgdal library.

#shapefile <- "~/GitHub/hw-1-musa5000/Lecture 1 - RegressionData.shp/RegressionData.shp"
shapefile <- "Lecture 1 - RegressionData.shp/RegressionData.shp"

c_map = read_sf(shapefile)

#names(c_map)

c_map <- c_map %>%
  mutate(LNMEDHVAL    = log(MEDHVAL),
         LNBELPOV100 = log(1 + NBelPov100),
         PCTVACANT, PCTSINGLES, PCTBACHMOR)

#maps the log of the dependent variable: LNMEDHVAL
ggplot(c_map) +
  geom_sf(aes(fill=LNMEDHVAL))+
  scale_fill_distiller(palette = "BuPu", direction = 1)+
  theme_void()+
  labs(title = "Log of Median House Value")

# maps 2-5: the four predictors
p1 <- ggplot(c_map) + geom_sf(aes(fill = LNBELPOV100)) +scale_fill_distiller(palette = "BuPu", direction = 1) + theme_void() + labs(title = "LNBELPOV100")
p2 <- ggplot(c_map) + geom_sf(aes(fill = PCTBACHMOR))   +scale_fill_distiller(palette = "BuPu", direction = 1) + theme_void() + labs(title = "PCTBACHMOR")
p3 <- ggplot(c_map) + geom_sf(aes(fill = PCTVACANT))    +scale_fill_distiller(palette = "BuPu", direction = 1) + theme_void() + labs(title = "PCTVACANT")
p4 <- ggplot(c_map) + geom_sf(aes(fill = PCTSINGLES))   +scale_fill_distiller(palette = "BuPu", direction = 1) + theme_void() +labs(title = "PCTSINGLES")

# combine into one 2x2 figure
(p1 | p2) / (p3 | p4)
```

## Stepwise Regression
# Sophia
```{r}
#| label: Using Stepwise regression to determine the best model based on the Akaike Information Criterion

# Full OLS model with all four predictors
model_full <- lm(
  LNMEDHVAL ~ PCTVACANT + PCTSINGLES + PCTBACHMOR + LNBELPOV100,
  data = regdata
)

summary(model_full)
```
```{r}
# Stepwise regression using AIC
step_model <- step(model_full)

step_model
```
```{r}
# View the sequence of models considered
step_model$anova
```
```{r}
# Show the final model selected by stepwise regression
formula(step_model)
```

## K-fold Cross Validation
# Sophia
```{r}
#| label: calculating RMSE and re-running the model with two predictors: PCTVACANT and MEDHHINC 

# 5-fold cross-validation for the full four-predictor model
cv_full <- CVlm(
  data = regdata,
  form.lm = model_full,
  m = 5,
  plotit = FALSE
)
```

```{r}
names(cv_full)
str(cv_full)
```

```{r}
# Calculate RMSE for the full model
rmse_full <- sqrt(
  mean((cv_full$LNMEDHVAL - cv_full$cvpred)^2, na.rm = TRUE)
)
rmse_full
```
```{r}
# Model with PCTVACANT and MEDHHINC
model_two <- lm(
  LNMEDHVAL ~ PCTVACANT + MEDHHINC,
  data = regdata
)
summary(model_two)

# 5-fold cross-validation
cv_two <- CVlm(
  data = regdata,
  form.lm = model_two,
  m = 5,
  plotit = FALSE
)

# Calculate 5-fold cross-validation RMSE
rmse_two <- sqrt(
  mean((cv_two$LNMEDHVAL - cv_two$cvpred)^2, na.rm = TRUE)
)

rmse_two
```
```{r}
rmse_results <- data.frame(
  Model = c(
    "Full model: 4 predictors",
    "Reduced model: PCTVACANT + MEDHHINC"
  ),
  RMSE = c(rmse_full, rmse_two)
)

rmse_results
```

# Results
## Standardized Regression Residuals
# Evelyn
```{r}
#| label: Histogram and Chloropleth map of standardized regression residuals


```

## Discussions and Limitations
# Evelyn
